---
title: "Problem set 2: More probability"
author: "(Your name here)"
subtitle: "Due October 9, 2023, at 9pm"
format: pdf
---

*\textsc{Note}: Start with the file `ps2_2023_more_probability.qmd` (available from the github repository at <https://github.com/UChicago-pol-methods/IntroQSS-F23/tree/main/assignments>). Modify that file to include your answers. Make sure you can "render" the file (e.g. in RStudio by clicking on the `Render` button). Submit both the qmd file and the PDF via Canvas.* 

## Problem 0: Bayes' Rule

One formulation of Bayes' Rule states that, if $\{A_1, A_2, A_3, \ldots \}$ is a partition of $\Omega$ and $B \in S$ with $P(B) > 0$, 

$$P(A_i \mid B) = \frac{P(B \mid A_i) P(A_i)}{\sum_i P(B \mid A_i) P(A_i)}$$

Here is a proof that is missing explanations for the steps: 

Step 1: 

$$P(A_i \cap B) =  P(A_i \mid B) P(B) = P(B \mid A_i) P(A_i)$$

Step 2: 

$$P(A_i \mid B) = \frac{P(B \mid A_i) P(A_i)}{P(B)}$$

Step 3: 

$$P(A_i \mid B) = \frac{P(B \mid A_i) P(A_i)}{\sum_i P(B \mid A_i) P(A_i)}$$


Explain each step in the proof: what definition(s)/rule(s)/law(s)/axiom(s)/condition(s)/mathematical operation(s) is the proof relying on?  

**Answer**: 


## Problem 1: Error rates in hypothesis testing

You have a fancy device that tests null hypotheses. Null hypotheses are statements about the world that can be either true or false. The device is designed to turn red when a null hypothesis is false and green when it is true, but it doesn't work perfectly: when a null hypothesis is false it turns red with probability 4/5 (i.e. it mistakenly turns green with probability 1/5), and when a null hypothesis is true it turns green with probability 19/20 (i.e. it mistakenly turns red with probability 1/20). Tests of different null hypotheses are independent, and 9/10 of the null hypotheses you test are true.


(1a) If you test 10 true null hypotheses in a row, what is the probability that the alarm turns red at least once? Explain your solution with reference to any axioms/definitions/rules/laws of probability you use.


**Answer**: 

(1b) Write a simulation to check your answer to (1a). That is, use `R` to generate many draws according to the random process described (testing ten true null hypotheses in a row), and confirm that the proportion of draws with at least one red light is approximately the same as in your answer above.

```{r}
# your code here 

```

(1c) What is the probability of getting a red light in any given test (i.e. when you don't know if the null hypothesis is true)? Explain your solution with reference to any axioms/definitions/rules/laws of probability you use. 

**Answer**: 


(1d) If the light turns red in a given test, what is the probability that the null hypothesis is false? Explain your solution with reference to any axioms/definitions/rules/laws of probability you use. 

**Answer**: 

(1e) Write a simulation to check your answer to (1c) and (1d). That is, use `R` to generate many draws according to the random process described (testing null hypotheses), and confirm your answer about the proportion of red-light-producing draws (1c) and the proportion of red-light-producing draws in which the null hypothesis is false (1d).

```{r}
# your code here

```


## Problem 2: discrete random variables

Consider tossing a coin four times. The results of the coin flips are assumed to be independent. Let $H$ and $T$ denote the outcome that it produces a head or a tail, respectively. 

(2a) Write the sample space $\Omega$ for this random process. 

**Answer**: 

\begin{align} 
\Omega &= \{ list, outcomes, \nonumber \\  
 & here \}  \nonumber
\end{align}

(2b) Are each of the outcomes in $\Omega$ equally likely to occur? Why or why not? 

**Answer**: 

(2c) Let the random variable $X$ denote the number of heads in four tosses of the coin. In each toss, the probability of getting a heads is $q$.  

Write the PMF $f(x)$ by replacing the "?"'s in the expression below. Use `R` to confirm that for a given value of $q$ (choose one!), $\sum_{x = 0}^4 f(x) = 1$.

**Answer**:

$$
f(x) = \begin{cases}
?  & x = 0 \\\
? & x = 1 \\\
? & x = 2 \\\
? & x = 3 \\\
? & x = 4  \\\
0 & \text{otherwise}
\end{cases}
$$


```{r}
# your code confirming that it sums to 1 for a value of q

```



(2d) Write the CDF $F(x)$ by replacing the "?"'s in the expression below.


**Answer**: 

$$
\Pr[X \leq x] = F(x) = \begin{cases}
?  & x <0 \\\
? & 0 \leq x < 1 \\\
? & 1 \leq x < 2 \\\
? & 2 \leq x < 3 \\\
? & 3 \leq x < 4 \\\
? & x \geq 4  \\\
\end{cases}
$$




(2e) Assume $q = 2/3$. Use the `sample()` function in `R` to draw a large number of samples from the PMF you specified above, and confirm that $F(3)$ agrees with your answer from (2d).

**Answer**:

```{r}
# your code here 

```



## Problem 3: continuous random variables 

(3a) Let $X$ be uniformly distributed between -5 and 3. Compute  $\Pr[X < 1]$ and $\Pr[-3 < X < 1/2]$ analytically (e.g. by computing the length and height of the area to be integrated) and confirm your results using `R`.

**Answer**: 

Confirmation in `R`:

```{r}
# your code here 
```


(3b) Let $X$ be normally distributed with mean 0 and standard deviation 1. Using `R`, compute $\Pr[-1.96 < X < 1.96]$.

**Answer**: 

```{r}
# your code here 
```

(3c) Let $X$ be normally distributed with mean 0 and standard deviation 2. Using `R`, use `dnorm()` to approximate $\Pr[-.1 < X < 0]$, and use `pnorm()` to see how close your approximation is. Explain why the two values should be close but not exactly the same.

**Answer**: 

```{r}
# your code here 
```



