---
title: "Problem set 3: Summarizing distributions"
author: "(Your name here)"
subtitle: "Due October 16, 2023, at 9pm"
format: pdf
---


<!-- Define special characters for operators -- leave this here -->
\def\E{{\textrm E}\,}
\def\V{{\textrm V}\,}



*\textsc{Note}: Start with the file `ps3_2023_summarizing_distributions.qmd` (available from the github repository at <https://github.com/UChicago-pol-methods/IntroQSS-F23/tree/main/assignments>). Modify that file to include your answers. Make sure you can "render" the file (e.g. in RStudio by clicking on the `Render` button). Submit both the qmd file and the PDF via Canvas.* 




## Question 1: Expected value 

Consider the random variable $X$ characterized by the following PMF.

| $x$  | $f(x)$ |
|:------:|:------------:|
|   -1   |     .3     |
|   2   |     .2     |
|   3   |     .5     |

(1a) Compute $\E[X]$. Show your work.

**Answer**:

$$
\begin{aligned} \E[X] &=  \end{aligned}  
$$
 
(1b) Write an `R` function to compute the expectation of any discrete random variable. The arguments to your function should include the values the random variable can take on (`x`) and the probability it takes on each value (`probs`). Use your function to confirm your answer from question (1a).

**Answer**:

```{r}
# your code here 
```

(1c) Compute the MSE ($\E[(X - c)^2]$) for $c = 0$ and $c = 1$. Show your work.

**Answer**:


(1d) Write a function to compute the MSE for any discrete random variable at a value `c`. The arguments to your function should include the values the random variable can take on (`x`), the probability it takes on each value (`probs`), and the value `c` being considered. Use your function to confirm your answers from (1c).

**Answer**:

```{r}
# your R code here 
```

(1e) Create a vector `cs` that contains numbers in the sequence `1.0, 1.1, 1.2, . . ., 3`. Using a for-loop and your function from (1d), compute the MSE for the random variable $X$ whose PMF was given above at each value of `cs` and store the result in a vector called `mses`.

**Answer**:

```{r}
# your R code here 
```

(1f) Using the `plot()` command, make a scatterplot showing the MSE of $X$ (vertical axis) at each value of $c \in \{1, 1.1, 1.2, \ldots, 3\}$ (horizontal axis). Use `abline()` to add a vertical line at $\E[X]$.

**Answer**:

```{r}
# your R code here 
```

## Question 2: Variance 


Consider the random variable $X$ characterized by the following PMF:

| $x$   | $f(x)$  |
|:-----:|:----------:|
|   0   |     .6     |
|   1   |     .1     |
|   2   |     .3     |

(2a) Confirm that the variance of $X$ is the same whether we compute it by the formula in Definition 2.1.12 or the Alternative Formula in Theorem 2.1.13. (We want the two variance computations, not the proof.)

**Answer**:


(2b) Write an `R` function to compute the variance of any discrete random variable. The arguments to your function should include the values the random variable can take on (`x`) and the probability it takes on each value (`probs`). Use your function to confirm your answer from question (2a).

**Answer**: 

```{r}
# your R code here

```

(2c) Explain definition/property/mathematical operation is being used in each step of the following proof.

For random variable $X$ and $c \in \mathbb{R}$, $V[X + c] = V[X]$,

\begin{align}
\V[X + c] &= \E\left[(X + c - \E[X + c])^2 \right] \tag{Step 1} \\
&= \E\left[(X + c - \E[X] - c)^2 \right]  \tag{Step 2} \\
&= \E\left[(X - \E[X])^2 \right]  \tag{Step 3} \\
&= \V[X] \tag{Step 4}
\end{align}


**Answer**: 

In Step 1, . . .

In Step 2, . . . 

In Step 3, . . . 

In Step 4, . . . 



(2d) Use your variance function to confirm that $\V[X + c] = \V[X] \forall c \in \mathbb{R}$. (You can just show this for one value of $c$.)

**Answer**:

```{r}
# your R code here 
```

(2e) Use your function to confirm that $V[aX] = a^2 V[X] \, \forall \, a \in \mathbb{R}$. (You can just show this for one value of $a$.)

**Answer**:

```{r}
# your R code here 
```




