---
title: "Problem set 3: Summarizing distributions"
author: "(Your name here)"
subtitle: "Due October 16, 2023, at 9pm"
format: pdf
---


<!-- Define special characters for operators -->
\def\E{{\textrm E}\,}
\def\V{{\textrm V}\,}



*\textsc{Note}: Start with the file `ps3_2023_summarizing_distributions.qmd` (available from the github repository at <https://github.com/UChicago-pol-methods/IntroQSS-F23/tree/main/assignments>). Modify that file to include your answers. Make sure you can "render" the file (e.g. in RStudio by clicking on the `Render` button). Submit both the qmd file and the PDF via Canvas.* 




## Question 1: Expected value 

Consider the random variable $X$ characterized by the following PMF.

| $x$  | $f(x)$ |
|:------:|:------------:|
|   -1   |     .3     |
|   2   |     .2     |
|   3   |     .5     |

(1a) Compute $\E[X]$. Show your work.

**Answer**:

$$
\begin{aligned} \E[X] &= \sum_x x f(x) \\
  &= -1 \times .3 + 2 \times .2 + 3 \times .5 \\
  &= 1.6 \end{aligned}  
$$
 
(1b) Write an `R` function to compute the expectation of any discrete random variable. The arguments to your function should include the values the random variable can take on (`x`) and the probability it takes on each value (`probs`). Use your function to confirm your answer from question (1a).

**Answer**:

```{r}
exp_func <- function(x, probs){
  sum(x * probs)
}
exp_func(x = c(-1, 2, 3), probs = c(.3, .2, .5))
```

(1c) Compute the MSE ($\E[(X - c)^2]$) for $c = 0$ and $c = 1$. Show your work.

**Answer**:

$$
\begin{aligned} \E[(X - 0)^2] &= \sum_x (x - 0)^2 f(x) \\
  &= (-1)^2 \times .3 + (2)^2 \times .2 + (3)^2 \times .5 \\
  &= 1\times .3 + 4 \times .2 + 9 \times .5 \\ 
  &= 5.6 \end{aligned}  
$$

$$
\begin{aligned} \E[(X - 1)^2] &= \sum_x (x - 1)^2 f(x) \\
  &= (-1 - 1)^2 \times .3 + (2 - 1)^2 \times .2 + (3 - 1)^2 \times .5 \\
  &= 4\times .3 + 1 \times .2 + 4 \times .5 \\ 
  &= 3.4 \end{aligned}  
$$

(1d) Write a function to compute the MSE for any discrete random variable at a value `c`. The arguments to your function should include the values the random variable can take on (`x`), the probability it takes on each value (`probs`), and the value `c` being considered. Use your function to confirm your answers from (1c).

**Answer**:

```{r}
mse_func <- function(x, probs, c){
  sum(probs*(x - c)^2)
}
mse_func(x = c(-1, 2, 3), probs = c(.3, .2, .5), c = 0)
mse_func(x = c(-1, 2, 3), probs = c(.3, .2, .5), c = 1)
```

(1e) Create a vector `cs` that contains numbers in the sequence `1.0, 1.1, 1.2, . . ., 3`. Using a for-loop and your function from (1d), compute the MSE for the random variable $X$ whose PMF was given above at each value of `cs` and store the result in a vector called `mses`.

**Answer**:

```{r}
cs <- seq(from = 1, to = 3, by = .1)
mses <- rep(NA, times = length(cs))
for(i in 1:length(mses)){
  mses[i] <- mse_func(x = c(-1, 2, 3), probs = c(.3, .2, .5), c = cs[i])
}
```

(1f) Using the `plot()` command, make a scatterplot showing the MSE of $X$ (vertical axis) at each value of $c \in \{1, 1.1, 1.2, \ldots, 3\}$ (horizontal axis). Use `abline()` to add a vertical line at $E[X]$.

**Answer**:

```{r}
plot(cs, mses)
abline(v = 1.6)
```

## Question 2: Variance 


Consider the random variable $X$ characterized by the following PMF:

| $x$   | $f(x)$  |
|:-----:|:----------:|
|   0   |     .6     |
|   1   |     .1     |
|   2   |     .3     |

(2a) Confirm that the variance of $X$ is the same whether we compute it by the formula in Definition 2.1.12 or the Alternative Formula in Theorem 2.1.13. (We want the two variance computations, not the proof.)

**Answer**:
First we calculate $\E[X]$: 

$$
\begin{aligned}
  \E[X] &= \sum_x x f(x)\\
  & = 0 \times .6 + 1 \times .1 + 2 \times .3 \\
  & = .7
\end{aligned}
$$

Using Definition 2.1.12:
$$
\begin{aligned} 
  \V[X] & = \E\left[(X - \E[X])^2\right] \\
  &=  \E\left[(X - .7)^2\right]\\
  &=  (-.7)^2 \times .6 + .3^2 \times .1 + 1.3^2 \times .3 \\
  &= .81  \end{aligned}  
$$

To use Theorem 2.1.13, we calculate $\E[X^2]$: 

$$
\begin{aligned}
  \E[X^2] &= \sum_x x^2 f(x)\\
  & = 0 \times .6 + 1 \times .1 + 4 \times .3 \\
  & = 1.3
\end{aligned}
$$

Using Theorem 2.1.13:
$$
\begin{aligned} 
  \V[X] & = \E[X^2]  - \E[X]^2 \\
  &=  1.3 - .7^2\\
  &=  .81 \end{aligned}  
$$

(2b) Write an `R` function to compute the variance of any discrete random variable. The arguments to your function should include the values the random variable can take on (`x`) and the probability it takes on each value (`probs`). Use your function to confirm your answer from question (2a).

**Answer**: 

```{r}
# two versions, both of which use the expectation function I wrote above 

# this version uses the E[(X - E[X])^2] formulation
var_func2 <- function(x, probs){
  ex <- exp_func(x, probs)
  exp_func((x - ex)^2, probs)
}

# this version uses the E[X^2] - E[X]^2 formulation
var_func <- function(x, probs){
  exp_func(x^2, probs) - exp_func(x, probs)^2
}

# confirming equality
var_func(x = c(0, 1, 2), probs = c(.6, .1, .3))
var_func2(x = c(0, 1, 2), probs = c(.6, .1, .3))
```

(2c) Explain what definition/property/mathematical operation is being used in each step of the following proof.

For random variable $X$ and $c \in \mathbb{R}$, $V[X + c] = V[X]$,

\begin{align}
\V[X + c] &= \E\left[(X + c - \E[X + c])^2 \right] \tag{Step 1} \\
&= \E\left[(X + c - \E[X] - c)^2 \right]  \tag{Step 2} \\
&= \E\left[(X - \E[X])^2 \right]  \tag{Step 3} \\
&= \V[X] \tag{Step 4}
\end{align}


**Answer**: 

In Step 1, . . .

In Step 2, . . . 

In Step 3, . . . 

In Step 4, . . . 




**Answer**: 

In Step 1, we use the definition of variance applied to the random variable $X + c$.  

In Step 2, we use the linearity of expectations and the fact that the expectation of a constant $c$ is $c$. 

In Step 3, we simplify by observing that $c - c = 0$.

In Step 4, we use the definition of variance.


(2d) Use your variance function to confirm that $\V[X + c] = \V[X] \forall c \in \mathbb{R}$. (You can just show this for one value of $c$.)

**Answer**

```{r}
var_func(x = c(0, 1, 2), probs = c(.6, .1, .3))
var_func(x = c(0, 1, 2) + 3, probs = c(.6, .1, .3))
```

(2e) Use your function to confirm that $V[aX] = a^2 V[X] \, \forall \, a \in \mathbb{R}$. (You can just show this for one value of $a$.)

**Answer**

```{r}
var_func(x = c(0, 1, 2), probs = c(.6, .1, .3))*9
var_func(x = c(0, 1, 2)*3, probs = c(.6, .1, .3))
```




