---
title: "Lab: Week 1"
author: "Moksha Sharma"
date: '2023-09-28'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


## Goals for today

In this lab, we will learn/review various basic yet essential operations in R. They will be helpful not only on the first problem set but also in projects you do in R in the coming years.

If you have not already done so, install the ```tidyverse``` package. Once it has been installed, you just need to load it at the start of every R session. 

## Vectors

A vector in R is one of the simplest data structures. It can be thought of as a list that contains the same type of elements. These elements can be numbers, logical values, characters, and even empty spaces (note: there are 6 categories of vectors in R that are described differently from the ones here, but it is not necessary to know what they are for our purposes). 

First, we will make a vector of our courses this quarter.

```{r, message = FALSE, echo = FALSE}

```

Now, make a different vector of the first five multiples of 2, starting with the number 2. 

```{r, message = FALSE, echo = FALSE}

```

Once a vector has been created, we can subset out specific elements using their position in the vector and ```[]```. For example, we can write a small line of code to add the first and third elements of the list.

```{r, message = FALSE, echo = FALSE}

```

This method of typing out the vector is not always the most convenient. Here, we only had to make a list of the first 5 multiples of 2. What if you had to list the first 10,000 multiples? How would you make that vector.

```{r, message = FALSE, echo = FALSE}

```

## Sampling

We will now learn how to create random samples in R. Since the samples are being generated (somewhat) randomly, we will all have different samples, unless we set a seed. Setting the seed produces the same random values each time, which is very important for producing reproducible results.

```{r, echo=FALSE, message=FALSE}
set.seed(60637)
```

The first task is to generate a sample of 50 numbers in the range 500-1000.

```{r, echo=FALSE, message=FALSE}

```

How many unique values are included in the sample? Can you think of any code that will count the number of unique values for you?

```{r, echo=FALSE, message=FALSE}

```

It seems that all values in the vector we just created are unique values. That is because one of the arguments of the ```sample``` function is "replace", and it is set to FALSE by default. This means that the sample is produced without replacement and thus there will be no repetation. You can change the argument to TRUE when you want to allow repetition. 

You can also specify the probability with which various elements will be drawn. Try to generate a sample of 50 numbers from the following normal distribution: $N(\mu=750, \sigma^2 = 10)$.

```{r, echo=FALSE, message=FALSE}

```

What would we expect the mean of the sample to be? Calculate it below:

```{r, echo=FALSE, message=FALSE}

```


## Working with Data

We will now use some of the skills we have learnt so far to add a new column to the "Iris" dataset. It is one of the built-in datasets in the ```tidyverse``` package and is commonly used for teaching demonstrations. 

```{r, message = FALSE, echo = FALSE}
# load the dataset

```

Let's assume that there were some gross measurement errors due to which, we now need to correct the values for the lengths and widths of the sepals. Without using functions from the ```tidyverse``` package, add a new column for sepal width that is the measured sepal width minus 1 unit. 

```{r, message = FALSE, echo = FALSE}

```

Do the same for the width column, but try to use a different function for adding the column.

Hint: we just used it.

```{r, message = FALSE, echo = FALSE}

```

Finally, let's add a column with a condition on the species. It should say "TRUE" if the species is "setosa" and "FALSE" otherwise.

```{r, message = FALSE, echo = FALSE}

```



